\documentclass{wg21}

\usepackage{xcolor}
\usepackage{soul}
\usepackage{ulem}
\usepackage{fullpage}
\usepackage{parskip}
\usepackage{csquotes}
\usepackage{listings}
\usepackage{minted}
\usepackage{enumitem}

\lstdefinestyle{base}{
  language=c++,
  breaklines=false,
  basicstyle=\ttfamily\color{black},
  moredelim=**[is][\color{green!50!black}]{@}{@},
  escapeinside={(*@}{@*)}
}

\newcommand{\cc}[1]{\mintinline{c++}{#1}}
\newminted[cpp]{c++}{}


\title{Wording for lambdas in unevaluated contexts}
\docnumber{P0315R3}
\audience{Core Working Group}
\author{Louis Dionne}{ldionne.2@gmail.com}
\authortwo{Hubert Tong}{hubert.reinterpretcast@gmail.com}

\begin{document}
\maketitle
\tableofcontents

\section{Revision history}
\begin{itemize}
  \item R0 -- Initial draft
  \item R1 -- Changed the wording to work around the resolution of \cite{DR1607},
              which conflicted with the initial wording. Also address the
              potential additional concerns raised by this wording change.
  \item R2 -- Applied feedback from EWG and CWG in Issaquah and CWG in Kona:
              \begin{itemize}
                \item a lambda expression is not part of the immediate context
                \item add a discussion about lambdas as non-type template arguments
                \item clarify the difference between the types of lambda expressions
                      in alias templates
                \item clarify ODR-equivalence of lambda expressions declared in
                      different TUs, and the meaning for function template
                      declarations
                \item clarify the situation for redeclarations of functions with
                      lambda-expressions
                \item rebase on top of the C++17 DIS
                \item editorial: extract the wording into its own section
              \end{itemize}
  \item R3 -- Applied changes per guidance from CWG in Toronto:
    \begin{itemize}
      \item editorial: hyphenate the term full-expression, italicize the
            term \textit{lambda-expression}, and remove sections useless
            for wording purposes
      \item replace wording in \textbf{[temp.over.link]} by equivalent but
            much simpler wording as suggested by Richard Smith
      \item add a discussion about implicit captures in \textit{lambda-expressions}
            in unevaluated contexts
    \end{itemize}
\end{itemize}

\section{Discussion on wording}
The core language changes introduced in this paper are a bit tricky. The reason
is that we remove many restrictions on lambda expressions, yet we still want to
keep closure types out of the signature of external functions, which would be
a nightmare for implementations. This discussion goes over all known possible
points of contingency to clarify them.

\begin{enumerate}
  \item \label{discussion.non-template}
    \hyperref[wording.non-template]{(wording)}
    With the removal of the restrictions on lambdas in unevaluated contexts,
    a concern is that lambda-expressions might then be able to appear in the
    signature of functions with external linkage, which would require implementations
    to generate a name for the associated closure type. However, since we wouldn't
    be able to attach to another ABI entity in some cases (such as lambda-expressions
    appearing at global scope), that would mean coming up with a mangling scheme
    that identifies the closure type from nothing but its own form. This, in turn,
    would require encoding its complete definition, which is burdensome for
    implementations and motivated the original restrictions on lambda-expressions.

    Fortunately, this specific problem can't arise in the case of non-template
    functions, even with the above removal of constraints on lambda-expressions.
    Indeed, according to \textbf{[basic.link] 6.5/8}, closure types have no linkage,
    and therefore they cannot appear in the signature of a function with external
    linkage (a function is a compound type):
    \begin{quote}
      \textit{[...]} A type is said to have linkage if and only if:\\
      \textit{[...]}\\
      - it is a compound type (6.9.2) other than a class or enumeration,
        compounded exclusively from types that have linkage; or \\
      \textit{[...]}
    \end{quote}

    However, to make it clear that closure types are never given a name for
    linkage purposes, we propose \hyperref[wording.non-template]{modifying
    \textbf{[dcl.typedef]}}.


  \item \label{discussion.template}
    \hyperref[wording.template]{(wording)}
    Another similar problem is that of lambda-expressions appearing in the
    signature of function templates. There are two ways this could happen.
    First, a lambda expression could appear not by itself, but indirectly, by
    being part of an expression which references a template parameter. Indeed,
    per \textbf{[temp.over.link] 17.5.6.1/4}:
    \begin{quote}
      When an expression that references a template parameter is used in the
      function parameter list or the return type in the declaration of a function
      template, the expression that references the template parameter is part of
      the signature of the function template.
    \end{quote}

    Thus, a function template declaration such as the following will require the
    implementation to make the lambda-expression part of the signature, which is
    specifically what we would like to avoid:
\begin{cpp}
template <int N>
void f(const char (*s)[([]{}, N)]) { }
\end{cpp}

    The other situation we would like to avoid is for implementations to have
    to figure out that the two following expressions are equivalent, either
    for linkage purposes (in different translation units) or for redeclaration
    purposes (in the same translation unit):
\begin{cpp}
template <int N> void f(const char (*s)[([]{ return N; })()]) { }
template <int N> void f(const char (*s)[([]{ return N; })()]) { }
\end{cpp}

    This case is slightly different from the first one, since the template
    parameter appears in the body of the lambda-expression, which is not
    considered part of the full-expression in the function parameter list.
    To make sure these cases do not happen, we propose
    \hyperref[wording.template]{amending \textbf{[temp.over.link]}}.


  \item \label{discussion.ODR}
    Another possible concern is the appearance of lambda-expressions in
    contexts that are constrained by the ODR. For example:

\begin{cpp}
// a.h:
template <typename T>
int counter() {
  static int cnt = 0;
  return cnt++;
}

inline int f() {
  return counter<decltype([] {})>();
}

// translation unit 1:
#include "a.h"
int foo() { return f(); }

// translation unit 2:
#include "a.h"
int bar() { return f(); }
\end{cpp}

    Given such code, a question might be whether \cc foo and \cc bar modify
    the same \cc cnt variable, since \cc f is defined in a header and it calls
    \cc counter with a closure type that is supposed to be unique. However,
    since \cc f is \cc inline, the resulting program is as-if there was a
    single definition of it, and so both functions end up modifying the same
    \cc cnt variable. This turns out not to be a problem for implementations,
    because they must already handle such cases where there is an ODR context
    to attach the closure type to. Thus, no wording change is required.


  \item \label{discussion.redeclarations}
    \hyperref[wording.redeclarations]{(wording)}
    A concern with allowing lambda-expressions in \textit{declarations} is
    that of dealing with entities that can be redeclared. It is already the
    case that no two lambda-expressions share the same closure type within a
    single translation unit:

\begin{cpp}
static decltype([] { }) f();
static decltype([] { }) f(); // invalid; return type mismatch

static decltype([] { }) g();
static decltype(g()) g(); // okay

static void h(decltype([] { }) *) { }
static void h(decltype([] { }) *) { }
h(nullptr); // ambiguous

using A = decltype([] { });
static void i(A *);
static void i(A *) { }
i(nullptr); // okay
\end{cpp}

   By further clarifying that the lambda-expressions in alias template
   specializations are unique to each specialization even if non-dependent,
   we conclude the following:

\begin{cpp}
template <typename T>
using B = decltype([] { });
static void j(B<char16_t> *) { }
static void j(B<char32_t> *) { }
j(nullptr); // ambiguous
\end{cpp}

    To make the above interpretation of the standard more obvious, we propose
    \hyperref[wording.redeclarations]{modifying \textbf{[temp.alias]}}.

    Furthermore, some questions were raised on the Core reflector regarding
    redeclarations like this:
\begin{cpp}
template <int N> static void k(decltype([]{ return 0; }()));
template <int N> static void k(decltype([]{ return 0; }())); // okay
template <int N> static void k(int);                         // okay
\end{cpp}

    These should be valid redeclarations, since the lambda expressions are
    evaluated, and they neither contain a template parameter in their body
    nor are part of a full-expression that contains one. Hence, the lambda-expression
    does not need to appear in the signature of the function, and the behavior
    is equivalent to this, without requiring any special wording:

\begin{cpp}
struct lambda { auto operator()() const { return 0; } };
template <int N> static void k(decltype(lambda{}()));
template <int N> static void k(decltype(lambda{}())); // okay today
template <int N> static void k(int);                  // okay today
\end{cpp}


  \item \label{discussion.sfinae}
    \hyperref[wording.sfinae]{(wording)}
    A concern with allowing lambda-expressions \textit{outside the body of
    the declaration of function templates} is the need to evaluate the validity
    of potentially complex expressions as part of template argument deduction.
    Indeed, without clarifying the wording, it is unclear whether implementations
    would be expected to support \textit{SFINAE} based on the validity of
    the body of a lambda-expression found in the declaration of a function
    template. Since this could be unwieldy for implementations, we choose not
    to require this in the current paper. Thus, if a lambda-expression appears
    inside the declaration of a function template and any part of it is ill-formed,
    then the program is ill-formed. To reflect this, we propose
    \hyperref[wording.sfinae]{adding a note at the end of \textbf{[temp.deduct]}}.


  \item \label{discussion.nontype}
    One usability question that has been raised with this paper is related
    to the usage of lambda expressions as non-type template arguments.

\begin{cpp}
// foo.h
template <auto> struct foo { };
foo<[]() {}> x;

// translation unit 1:
#include "foo.h"

// translation unit 2:
#include "foo.h"
\end{cpp}

    With the current wording that would be an ODR violation. Is this something
    we want to do something about?


  \item \label{discussion.capture}
    Whether a \textit{lambda-expression} implicitly captures an entity
    currently depends on whether the \textit{lambda-expression}
    (\textbf{[expr.prim.lambda.capture] 8.1.5.2/7}):
    \begin{enumerate}
      \item odr-uses the entity (in the case of a variable),
      \item odr-uses \cc{this} (in the case of the object designated by \cc{*this}), or
      \item names the entity in a potentially-evaluated expression where the
            enclosing full-expression depends on a generic lambda parameter
            declared within the reaching scope of the \textit{lambda-expression}.
    \end{enumerate}

    Since an unevaluated context does not satisfy the above criteria, there
    was a question regarding whether variables could be implicitly captured
    by a \textit{lambda-expression} appearing in an unevaluated context:
\begin{cpp}
void f(int i) {
  auto lambda = [=]{ return i; }; // captures i
  static_assert(sizeof(L) == sizeof([=]{ return i; })); // Error, i was not captured?
}
\end{cpp}

    The desire is that entities be implicitly captured by \textit{lambda-expressions}
    even when the lambda appears in an unevaluated context. However, according
    to the great oracle of C++, Richard Smith, this already works as desired:

    \begin{quote}
      Lambdas in unevaluated operands have a lot more value if we also permit
      default initialization of non-capturing lambdas (that gives us the ability
      to use \cc{std::set<T, decltype([](T a, T b) { return ... })>}, for
      instance). If we *do* permit that, then we should treat the body of at
      least a non-capturing lambda as *not* being an unevaluated operand even
      when the \textit{lambda-expression} is lexically within one. That's
      actually already the result that our current definition of "subexpression"
      gives. And that in turn means that lambdas would need to capture
      enclosing variables that they use, even in unevaluated operands.
    \end{quote}

    Hence, no wording change is required.

  \item \label{discussion.blockscope}
    Consider the following example:

\begin{cpp}
auto f(int i) -> decltype([=](auto g) { return g(i); }) { }
\end{cpp}

    As Richard Smith says:
    \begin{quote}
      It's not completely clear whether this would be valid: the
      \textit{lambda-expression} is not lexically within a block scope,
      but names from the function's block scope are visible at the point of
      the \textit{lambda-expression}, so is this a local \textit{lambda-expression}
      or not? If \cc{f} were a non-defining function declaration, I think
      \textbf{[basic.scope.proto]} makes it clear that the lambda would not
      be a local \textit{lambda-expression} and \textbf{[expr.prim.lambda.capture] 8.1.5.2/3}
      would disallow it having any captures. It'd be good to clear up exactly
      what it means for the smallest enclosing scope of a \textit{lambda-expression}
      to be a block scope; I don't think this mattered up until now.
    \end{quote}

    For the time being, and for the purpose of making this paper progress, we
    do not tackle this limitation and leave it ill-formed to perform such
    captures.
\end{enumerate}


\section{Proposed Wording}
The wording is based on the C++17 DIS \cite{N4659}:

\begin{enumerate}
  \item \label{wording.remove}
    In \textbf{[expr.prim.lambda] 8.1.5/2}:
    \begin{quote}
      The evaluation of a \textit{lambda-expression} results in a prvalue
      temporary (12.2). This temporary is called the closure object. \removed{A
      \textit{lambda-expression} shall not appear in an unevaluated operand
      (Clause 8), in a \textit{template-argument}, in an \textit{alias-declaration},
      in a typedef declaration, or in the declaration of a function or function
      template outside its function body and default arguments. [ Note: The
      intention is to prevent lambdas from appearing in a signature.
      -- end note ]} [ Note: A closure object behaves like a function object
      (23.14). -- end note]
    \end{quote}

  \item \label{wording.non-template} \hyperref[discussion.non-template]{(discussion)}
    In \textbf{[dcl.typedef] 10.1.3/9}:
    \begin{quote}
      If the typedef declaration defines an unnamed class (or enum), the first
      typedef-name declared by the declaration to be that class type (or enum type)
      is used to denote the class type (or enum type) for linkage purposes only (6.5).
      \added{However, a closure type is never given a name for linkage purposes.}
      [ \textit{Example:}
\begin{codeblock}
typedef struct { } *ps, S; // S is the class name for linkage purposes
@\added{typedef decltype([]\{\}) C; // the closure type has no name for linkage purposes}@
\end{codeblock}
      \textit{-- end example} ]
    \end{quote}
  \item \label{wording.template} \hyperref[discussion.template]{(discussion)}
    In \textbf{[temp.over.link] 17.5.6.1/5}:
    \begin{quote}
      Two expressions involving template parameters are considered \textit{equivalent}
      if two function definitions containing the expressions would satisfy the
      one-definition rule (6.2), except that the tokens used to name the template
      parameters may differ as long as a token used to name a template parameter
      in one expression is replaced by another token that names the same template
      parameter in the other expression. \added{Two \textit{lambda-expressions}
      are never considered equivalent. [ \textit{Note:} The intent is to avoid
      \textit{lambda-expressions} appearing in the signature of a function
      template with external linkage. \textit{-- end note} ]}
    \end{quote}

    Also, add the following example after \textbf{[temp.over.link] 17.5.6.1/5}:
\begin{codeblock}
@\added{
// ill-formed, no diagnostic required: the two expressions are\\
// functionally equivalent but not equivalent\\
template <int N> void foo(const char (*s)[([]\{\}, N)]); // TU 1\\
template <int N> void foo(const char (*s)[([]\{\}, N)]); // TU 2\\
\\
// ill-formed, no diagnostic required: the two lambda-expressions are\\
// functionally equivalent but not equivalent\\
template <int N> void bar(const char (*s)[([]() \{ return N; \})()]); // TU 1\\
template <int N> void bar(const char (*s)[([]() \{ return N; \})()]); // TU 2\\
\\
// two different declarations because the non-dependent portions differ\\
template <class T> void baz(T, char (*s)[sizeof(decltype([]\{ return 0; \}))]);\\
template <class T> void baz(T, char (*s)[sizeof(decltype([]\{ return 0; \}))]);\\
\\
// two different declarations because the non-dependent portions differ\\
template <class T> void spam(decltype([]\{\}) (*s)[sizeof(T)]);\\
template <class T> void spam(decltype([]\{\}) (*s)[sizeof(T)]);\\
}@
\end{codeblock}

  \item \label{wording.redeclarations} \hyperref[discussion.redeclarations]{(discussion)}
    Add the following paragraph at the end of \textbf{[temp.alias] 17.5.7}:
    \begin{quote}
      \added{
      The type of a lambda expression appearing in an alias template declaration
      is different between instantiations of that template, even when the lambda
      expression is not dependent.
      [ \textit{Example:}
      }
\begin{codeblock}
@\added{
template <class T>\\
using A = decltype([] \{ \});\\
// A<int> and A<char> refer to different closure types
}@
\end{codeblock}
      \added{\textit{-- end example} ]}
    \end{quote}

  \item \label{wording.sfinae} \hyperref[discussion.sfinae]{(discussion)}
    Add the following after \textbf{[temp.deduct] 17.8.2/8} (note that the
    term \textit{immediate context} is not defined formally in the standard,
    which is the subject of \cite{CWG1844}):
    \begin{quote}
      \added{
      A lambda expression appearing in a function type or a template parameter
      is not considered part of the immediate context for the purposes of
      template argument deduction. [ \textit{Note:}  The intent is to avoid
      requiring implementations to deal with substitution failure involving
      arbitrary statements. [ \textit{Example:}
      }
\begin{codeblock}
@\added{
template <class T>\\
auto f(T) -> decltype([]() \{ T::invalid; \} ());\\
void f(...);\\
f(0); // error: invalid expression not part of the immediate context\\
\\
template <class T, std::size_t = sizeof([]() \{ T::invalid; \})>\\
void g(T);\\
void g(...);\\
g(0); // error: invalid expression not part of the immediate context\\
\\
template <class T>\\
auto h(T) -> decltype([x = T::invalid]() \{ \});\\
void h(...);\\
h(0); // error: invalid expression not part of the immediate context\\
\\
template <class T>\\
auto i(T) -> decltype([]() -> typename T::invalid \{ \});\\
void i(...);\\
i(0); // error: invalid expression not part of the immediate context\\
\\
template <class T>\\
auto j(T t) -> decltype([](auto x) -> decltype(x.invalid) \{ \} (t));\\
void j(...);\\
j(0); // deduction fails on \#1, calls \#2
}@
\end{codeblock}
      \added{
      \textit{-- end example} ]
      \textit{-- end note} ]
      }
    \end{quote}

  %
  % NOTE: THIS CAN BE REMOVED ONCE I'M SURE NO WORDING IS REQUIRED FOR THIS
  %
  % \item \label{wording.capture} \hyperref[discussion.capture]{(discussion)}
  %   Add in \textbf{[expr.prim.lambda.capture] 8.1.5.2/7}:
  %   \begin{quote}
  %     A \textit{lambda-expression} with an associated \textit{capture-default}
  %     that does not explicitly capture \cc{*this} or a variable with automatic
  %     storage duration (this excludes any \textit{id-expression} that has been
  %     found to refer to an \textit{init-capture}'s associated non-static data
  %     member), is said to implicitly capture the entity (i.e., \cc{*this} or a
  %     variable) if the \textit{compound-statement}:
  %     \begin{itemize}
  %       \item odr-uses (6.2) the entity (in the case of a variable),
  %       \item odr-uses (6.2) \cc{this} (in the case of the object designated
  %             by \cc{*this}), or
  %       \item names the entity in a potentially evaluated expression where the
  %             enclosing full-expression depends on a generic lambda parameter
  %             declared within the reaching scope of the \textit{lambda-expression}.
  %     \end{itemize}
  %     [ \textit{Example:}
  %     \\ \dots \\
  %     \textit{-- end example} ] All such implicitly captured entities shall be
  %     declared within the reaching scope of the lambda expression. \added{For
  %     the purpose of determining implicit captures, the \textit{compound-statement}
  %     of a \textit{lambda-expression} is an evaluated context.}
  %     \begin{note}The implicit capture of an entity by a nested
  %     \textit{lambda-expression} can cause its implicit capture
  %     by the containing \textit{lambda-expression} (see below).
  %     Implicit odr-uses of \cc{this} can result in implicit capture.
  %     \end{note}
  %   \end{quote}
\end{enumerate}


\section{References}
\renewcommand{\section}[2]{}%
\begin{thebibliography}{9}

  \bibitem[N4659]{N4659}
    Richard Smith,
    \emph{Working Draft, Standard for Programming Language C++}\newline
    \url{http://www.open-std.org/jtc1/sc22/wg21/docs/papers/2017/n4659.pdf}

  \bibitem[DR1607]{DR1607}
    Daniel Kr{\"u}gler,
    \emph{Lambdas in template parameters}\newline
    \url{http://www.open-std.org/jtc1/sc22/wg21/docs/cwg_defects.html#1607}

  \bibitem[CWG1844]{CWG1844}
    Richard Smith,
    \emph{Defining "immediate context"}\newline
    \url{http://www.open-std.org/jtc1/sc22/wg21/docs/cwg_active.html#1844}

\end{thebibliography}

\end{document}
