\documentclass{wg21}

\usepackage{xcolor}
\usepackage{soul}
\usepackage{ulem}
\usepackage{fullpage}
\usepackage{parskip}
\usepackage{csquotes}
\usepackage{listings}
\usepackage{minted}
\usepackage{enumitem}

\lstdefinestyle{base}{
  language=c++,
  breaklines=false,
  basicstyle=\ttfamily\color{black},
  moredelim=**[is][\color{green!50!black}]{@}{@},
  escapeinside={(*@}{@*)}
}

\newcommand{\cc}[1]{\mintinline{c++}{#1}}
\newminted[cpp]{c++}{}


\title{Default constructible and assignable stateless lambdas}
\docnumber{D0624R2}
\audience{Evolution Working Group}
\author{Louis Dionne}{ldionne.2@gmail.com}

\begin{document}
\maketitle

\section{Revision history}
\begin{itemize}
  \item R0 -- Initial draft
  \item R1 \begin{itemize}
             \item Rebase wording on top of the current working draft N4659.
             \item Make stateless lambdas assignable per EWG direction in Toronto.
           \end{itemize}
  \item R2 \begin{itemize}
             \item Rebase wording on top of current working draft \cite{N4700} (no-op).
             \item Fix wording to keep the assignment operator deleted for stateful lambdas.
             \item Record the intent that construction and assignment operations be trivial.
           \end{itemize}
\end{itemize}


\section{Introduction}
Lambda expressions are a very convenient way of creating inline function objects,
especially for passing to higher order algorithms. Furthermore, the Standard
treats stateless lambdas specially, giving them an implicit conversion to a
corresponding function pointer, which is incredibly useful for storing these
in data structures. Unfortunately, stateless lambdas are neither default
constructible nor assignable, which precludes many interesting use cases.
This paper proposes making stateless lambdas default constructible and
assignable, so that code like this is valid:

\begin{cpp}
// in library.hpp
auto greater = [](auto x, auto y) { return x > y; };

// in user.cpp
// No need to care whether `greater` is a lambda or a function object
std::map<std::string, int, decltype(greater)> map;
\end{cpp}


\section{Motivation}
The first argument for this change is consistency. Indeed, since a lambda is
just syntactic sugar for a function object, it makes sense that the two can be
used interchangeably as much as possible:

\begin{cpp}
auto greater = [](auto x, auto y) { return x > y; };
std::map<std::string, int, decltype(greater)> map;

// should be the same as

struct {
  template <typename X, typename Y>
  auto operator()(X x, Y y) const { return x > y; }
} greater;
std::map<std::string, int, decltype(greater)> map;
\end{cpp}

Instead, the former produces an error message saying something about the use
of the deleted default constructor of the lambda used inside the map. Furthermore,
since lambdas are currently not assignable, the following would also produce an
error:

\begin{cpp}
auto greater = [](auto x, auto y) { return x > y; };
std::map<std::string, int, decltype(greater)> map1{...}, map2{...};
map1 = map2; // oops!
\end{cpp}

Apart from consistency, the second argument is that making stateless lambdas
more like normal function objects opens a very interesting design space for
writing libraries. An important tenet of C++ is to give power to experts
(while keeping simple things simple), and this tiny feature opens up a whole
new class of use cases. For example, default-constructible lambdas allow the
\cite{Dyno} library to implement runtime polymorphism with value semantics
(like \cc{std::function}, but for arbitrary interfaces) at the library level:

\begin{cpp}
// Define the interface of something that can be drawn
struct Drawable : decltype(dyno::requires(
  "draw"_s = dyno::function<void (dyno::T const&, std::ostream&)>
)) { };

// Define how a type can satisfy the above concept
template <typename T>
auto const dyno::concept_map<Drawable, T> = dyno::make_concept_map(
  "draw"_s = [](T const& self, std::ostream& out) { self.draw(out); }
);

// Define an object that can hold anything that can be drawn
struct drawable {
  // ... boilerplate that would go away with reflection ...
};

struct Square {
  void draw(std::ostream& out) const { out << "Square"; }
};

struct Circle {
  void draw(std::ostream& out) const { out << "Circle"; }
};

void f(drawable d) {
  d.draw(std::cout);
}

// drawable is like std::function, but for Drawables instead of Callables
f(Square{...});
f(Circle{...});
\end{cpp}

While both the purpose of the library and its implementation are completely
beyond the scope of this paper, the idea is that we need to transport the
type of the lambda around without an instance of it, and then default-construct
that lambda to fill an entry in the virtual table of an interface. This is one
use case, and there are possibly many more to come, that currently require a
gruesome hack to work (see \href{https://github.com/ldionne/dyno/blob/03eaeded898225660787f03655edb89642a72e7c/include/dyno/detail/empty_object.hpp#L13}{this Gist}).


\section{Proposed Wording}
The wording is based on the working paper \cite{N4700}. Replace in
\textbf{[expr.prim.lambda] 8.1.5.1/11}:

\begin{quote}
  \removed{The closure type associated with a \textit{lambda-expression} has
  no default constructor and a deleted copy assignment operator.}
  \added{The closure type associated with a \textit{lambda-expression} has a
  deleted default constructor if the \textit{lambda-expression} has a
  \textit{lambda-capture}, and a defaulted default constructor otherwise.
  It has a deleted copy assignment operator if the \textit{lambda-expression}
  has a \textit{lambda-capture}, and a defaulted copy assignment operator otherwise.}
  It has a defaulted copy constructor and a defaulted move constructor (15.8).
  \begin{note}
  These special member functions are implicitly defined as usual, and might
  therefore be defined as deleted.
  \end{note}
\end{quote}

When discussed in EWG, interest in having stateless lambdas be \textit{trivially}
default constructible was expressed, if CWG thought it to be a good idea. The
current wording may not achieve that, so minor tweaking might be necessary if
trivial default constructibility is desired.


\section{References}
\renewcommand{\section}[2]{}%
\begin{thebibliography}{9}

  \bibitem[Dyno]{Dyno}
    Louis Dionne,
    \emph{Dyno: Runtime polymorphism done right}\newline
    \url{https://github.com/ldionne/dyno}

  \bibitem[N4700]{N4700}
    Richard Smith,
    \emph{Working Draft, Standard for Programming Language C++}\newline
    \url{http://www.open-std.org/jtc1/sc22/wg21/docs/papers/2017/n4700.pdf}

\end{thebibliography}

\end{document}
